JD notes on effects

### 2022-06-18 manuscript/abstract.tex
Heading to school

In applications that use generalized linear (mixed) models, outcome predictions are often of interest. For models that involve complex multiplicative interactions, additional non-focal input variables or nonlinear link functions, the estimated coefficients are not readily interpretable. A general way to summarize these kind of models is through predictions, that depend not only on choice of representative values of focal input variables but are also sensitive to which values of non-focal input variables are chosen. The most common approach is generating predictions at a reference point, usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an average case in the population. In the presence of additional non-focal input variables, nonlinear link functions, random effect terms, etc., mean-based approach generate predictions that are biased and may not be consistent with the observed quantities. An alternative is the whole-sample-based approach which estimates the average effect in the population. Moreover, anchored effects provide an alternative and a more clear way to describe uncertainty associated with the focal input variable. In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction. We also present an alternative way, anchored confidence intervals, to describe uncertainties associated with these predictions.

>>>>

In applications that use generalized linear (mixed) models, outcome predictions are often of interest. For models that involve complex multiplicative interactions, additional non-focal input variables or nonlinear link functions, the estimated coefficients are not readily interpretable. A general way to summarize these kind of models is through predictions, that depend not only on choice of representative values of focal input variables but are also sensitive to which values of non-focal input variables are chosen. The most common approach is generating predictions at a reference point, usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an average case in the population. In the presence of additional non-focal input variables, nonlinear link functions, random effect terms, etc., mean-based approach generate predictions that are biased and may not be consistent with the observed quantities. An alternative is the whole-sample-based approach which estimates the average effect in the population. Moreover, effects confidence intervals provide an alternative and a more clear way to describe uncertainty associated with the focal input variable. In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction. We also present an alternative way, effects confidence intervals, to describe uncertainties associated with these predictions.

### 2022-06-18 manuscript/body.tex
Heading to school

In applications that use regression models, outcome predictions are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models interactions or other nonlinear transformations, the magnitude of the effect of change in the outcome depends on the values of the input variable of interest and the other remaining input variables. Hence, the conclusions one can make about the estimated predictions greatly depends on how well we choose the values of the other input variables, i.e., \emph{non-focal} input variables.

@@

Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and input variables e.g., interactions between and nonlinear transformation on the input variables, (and via splines for example), and nonlinear transformations via link functions. This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}. Further, coefficient estimates of GLM models involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model. An alternative is to explore \emph{predictions} and \emph{effects} \citep{fox2009effect, leeper2017package, lenth2018package}.  For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.

@@

When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction terms or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret. Generating these quantities together with the associated confidence intervals for regression models has a number of challenges. In particular:
\begin{enumerate}
\item choice of representative values of focal input variable and the \emph{reference point} for non-focal input variables, especially in multivariate models
\item  uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal input variables 
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
\end{enumerate}

The most common way of choosing the representative values is taking unique levels of the focal input variable if discrete or taking appropriately sized quantiles if continuous, and then calculating the predictions while holding non-focal (other input variables other than the focal input variable) input variables at their reference point (values we chose for the non-focal input variables e.g., means) \citep{fox2009effect, hanmer2013behind}. This procedure generates -- predictions (at least in our case), \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the input variable, for an ``average case''. More specifically, the basic idea is to compute the expected outcome by holding the non-focal input variables constant, for example at their mean, while varying the focal input variable, with the goal that estimates represent how the model responds to the changes in the focal input variable. As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.

The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively. However, there are a number of choices one can make when considering this approach -- for example, in the presence of interaction, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}). For simple models (with no interactions), the two approaches give the same predictions. However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:

@@

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$ $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.

@@

In the absence of interaction (model~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A. However, in the presence of interaction, even as simple as the one in model~\ref{eq:simple_inter_higher}, the estimates starts to differ. In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B. To generate Fig~\ref{fig:justify_plots}A, all the three packages average $x_2$ and $x_3$, i.e., the input variables. On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$). In particular \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$. In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.

One may be interested in the uncertainties associated with the focal input variable of interest only, excluding other uncertainties due to other non-focal input variables -- effects. However, currently, the two packages do not provide straightforward way do achieve this. To illustrate this, we generate predictions of $x_2$ from model~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence bands, as shown in Fig~\ref{fig:justify_ci_plots}.

\begin{figure}
\begin{center}
\includegraphics{justify_preds.isolate.pdf}
\end{center}
\caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above. The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predcitions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal input variable, i.e., the model center, corresponds to the effects of the focal input variable from \pkg{varpred}. For simple models, the isolated confidence bands crosses at the model center. The wider curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables. On the other hand, effects (narrower crossing lines) only take into account the main effect uncertainty of the focal input variable.}
\label{fig:justify_ci_plots}
\end{figure}

@@

Using simple indirect mediation simulation model described in \nameref{S3_Appendix}, fitted two models -- \emph{non-mediated} which models \emph{z} as a function of \emph{x} (univariate model) and the \emph{mediated} which models \emph{z} as a function of both \emph{x} and \emph{y} (multivariate model). In both cases, we compared mean-based and the whole-sample-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.

\begin{figure}
\begin{center}
\includegraphics{mediate_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models. A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal predictors) sources of bias. Consequently, for both approaches, the average predicted probability is very close to the observed proportion. B: When the mediator variable is included, there is no direct effect of $x$ on $z$ and as a result the predictions do not necessarily align with the observations. However, the whole-sample-based approach still closely approximates the average proportion in the simulated data. The horizontal blue, black and the red dotted lines are the average mean-based, whole-sample-based predicted probability and observed proportion, respectively. The vertical dotted black line represents the mean of the focal predictor, and the point at which it crosses the red dotted line represents the expected ``perfect'' prediction at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

From Fig~\ref{fig:pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$. By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches. However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}. Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $y$ and as a result, our prediction do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).

@@

Our simulations examples majorly focused on simple linear and logistic models due their wide range of usage and application. In addition, these models act as a starting point for building other complex models, including mixed effect models and models with categorical predictors. However, the logic for extension of approaches to more complex models, including other forms of nonlinear link functions is very straightforward. In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework, including multivariate binary outcome models.

@@

Generating outcome predictions or predicted probabilities from simple and generalized linear (mixed) models is not only important but also, generating quantities which are consistent with the observed values should be of interest. However, many studies still report coefficient estimates from generalized models like probit, logistic, etc., \citep{hanmer2013behind}, which are subject of less clarity due to lack of direct link to the outcome of interest.

The argument and results we present in this paper supports a greater need for a shift on focus on what and how to present predictions from generalized models. For example, we believe that effect-styled confidence intervals could provide more clarity concerning the uncertainty due to the predictor of interests as opposed to the conventional way of incorporating everything.

@@

\noindent\textbf{Supervision:} Jonathan Dushoff, Benjamin M. Bolker

>>>>

In applications that use regression models, outcome predictions are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models with complex interactions, the magnitude of the effect of change in the outcome depends on the values of the input variable of interest and the other input variables, i.e., \emph{non-focal} input variables. In other words, the conclusions we make about the predictions greatly depends on how well we choose the values of the non-focal input variables.

@@

Simple OLS regression models are flexible and can easily be generalized, as generalized linear (mixed) models (GL(M)Ms), to examine more complex relationships, including nonlinear relationships between response and input variables e.g., higher order interactions via polynomials and splines, nonlinear transformation on the input variables, and nonlinear transformations via link functions. This flexibility comes with additional complications, for example, complex multivariate models may risk misinterpretation, and miscalculation of quantities of interest \citep{berry_improving_2012, leeper2017interpreting}. Further, coefficient estimates of GLMs involving nonlinear link functions lack direct interpretation\citep{leeper2017interpreting}, meaning that interpretation of derived quantities from these estimates requires some understanding of the specified model. An alternative is to explore \emph{predictions} and \emph{effects} \citep{fox2009effect, leeper2017package, lenth2018package}.  For example, as a way to plan, public health officials may want to know the effects of household \code{income}, \code{wealthindex}, etc, on the predicted probability that slum dwellers will have improved water services.

@@

When visually presented, predictions and effects provide a unified and intuitive way of describing relationships from a fitted model, especially complex models involving interaction or some kind of transformations on the input variables whose estimates are usually, but not always, harder to interpret. Generating these quantities together with the associated confidence intervals for regression models has a number of challenges. In particular:
\begin{enumerate}
\item choice of representative values of focal input variable and the \emph{reference point} for non-focal input variables, especially in multivariate models
\item  uncertainty estimation -- appropriate choice of \emph{anchor} for computing confidence intervals; and how to incorporate the uncertainty due to non-focal input variables 
\item bias in the expected mean prediction induced by the nonlinear transformation of the response variable (especially in GL(M)Ms)
\end{enumerate}

The most common way of choosing the representative values is taking unique levels of the focal input variable if discrete or taking appropriately sized quantiles if continuous. The predictions are then calculated by holding the non-focal input variables at their reference point (values chosen for the non-focal input variables, e.g., means), while varying the focal input variable, with the goal that the estimates represent how the model responds to the changes in the focal input variable \citep{fox2009effect, hanmer2013behind}. This procedure generates -- \emph{predictor effects} \citep{fox2009effect}, \emph{marginal predictions} \citep{leeper2017package} or \emph{estimated marginal means} \citep{lenth2018package}. In this article, we refer to these quantities as predictions and effects since they should, for example, tell us what we would expect the response to be at a particular value or level of the focal input variable, for an ``average case''. . As mentioned above, in our case, the key difference between predictions and effects is in the computation of the confidence intervals.

The most commonly used \proglang{R} software packages for generating predictions (\pkg{emmeans} and \pkg{effects}), by default, use the average of \emph{input} and \emph{predictor variables} as the reference point for continuous and categorical input variables, respectively. However, there are a number of choices one can make when generating predictions -- for example, in the presence of interactions, averaging the interactions (averaging the product of interacting predictor variables) versus taking the product of the averages of the interacting predictor variables (the default for \pkg{emmeans} and \pkg{effects}). For simple models (with no interactions), the two approaches give the same predictions. However, for complex models, we demonstrate that the two approaches can produce substantially different results and show that averaging the interactions closely matches the observed values. To illustrate this, consider models~\ref{eq:simple_inter_higher_no_interaction} and \ref{eq:simple_inter_higher} below, with $x_1$ as the focal input variable:

@@

We first simulate data with the two models, such that $x_{1,2,3} \sim \mathrm{Normal}(0, 1)$, $\beta_0 = 5$, $\beta_1 = -3$, $\beta_2 = 1$, $\beta_3 = 2$, $\beta_{23} = 5$, $\epsilon \sim \mathrm{Normal}(0, \sigma^2)$ and $\sigma^2 = 1$, and then compare the predictions from the \pkg{emmeans}, \pkg{effects} and our proposed alternative (\pkg{varpred}) to the ``true'' predictions (predictions based on the true simulation parameters) and observed average i.e., $\bar{y}$, as shown in Fig~\ref{fig:justify_plots}.

@@

In the absence of interaction (Equation~\ref{eq:simple_inter_higher_no_interaction}), the three approaches produce the same estimates, which match the simulated values, Fig~\ref{fig:justify_plots}A. However, in the presence of interaction, even as simple as the one in Equation~\ref{eq:simple_inter_higher}, the estimates starts to differ. In particular, \pkg{emmeans} and \pkg{effects} give similar estimates ($\bar{\hat{y}}$) but different from the \pkg{varpred}'s which, however, is very close to the simulated average ($\bar{y}$), Fig~\ref{fig:justify_plots}B. To generate Fig~\ref{fig:justify_plots}A,  \pkg{emmeans} and \pkg{effects} average $x_2$ and $x_3$, i.e., the input variables while \pkg{varpred} averages the predictor variables corresponds to $x_2$ and $x_3$. On other hand, to generate Fig~\ref{fig:justify_plots}B, the difference in the estimates lies on how each of the packages average the interaction term ($x_2x_3$). In particular \pkg{emmeans} and \pkg{effects} use input variables to compute $\bar{x_2}\bar{x_3}$ while \pkg{varpred} uses predictor variables to compute $\over{x_2x_3}$. In the case of Fig~\ref{fig:justify_plots}A, averaging the input variables is equivalent to averaging the predictor variables.

One may be interested in the uncertainties associated with the focal input variable only, excluding other uncertainties due to other non-focal input variables -- effects. However, currently, the two packages do not provide a straightforward way do achieve this. To illustrate this, we generate predictions of $x_2$ from Equation~\ref{eq:simple_inter_higher_no_interaction} together with the associated $95\%$ confidence intervals, as shown in Fig~\ref{fig:justify_ci_plots}.

\begin{figure}
\begin{center}
\includegraphics{justify_preds.isolate.pdf}
\end{center}
\caption{{\bf A comparison of conventional $95\%$ confidence bands and effects.} The description of horizontal, vertical and trend lines remain the same as above. The wider dotted blue curves overlaying the green curves corresponds to the conventional confidence bands around the predictions from \pkg{emmeans} and \pkg{effects}, while the narrower black curves crossing at the mean of the focal input variable, i.e., the model center, corresponds to the effects of the focal input variable from \pkg{varpred}. For simple models, the isolated confidence bands crosses at the model center. The wider curves incorporates not only the uncertainties due to intercept term but also other non-focal input variables. On the other hand, effects (narrower crossing lines) only take into account the main effect uncertainty of the focal input variable.}
\label{fig:justify_ci_plots}
\end{figure}

@@

A \emph{mediated} effect is a situation in which only some of the input variables have a \emph{causal} impact on the outcome, even though all the variables are (somehow) associated with the outcome. For example, the age at marriage may be strongly associated with divorce rate but only married people divorce. So does marriage \emph{cause} divorce?

For simplicity, we consider an hypothetical simulation for three variables: predictors $x$ and $y$; and outcome variable $z$; such that
%
\begin{center}
\begin{tikzpicture}
    \node (1) at (0,0) {x};
    \node (2) [right = of 1] {y};
    \node (3) [below = of 2] {z};
    \path (1) edge  (2);
    \path (1) edge (2);
    \path (2) edge (3);
    \path (1) edge (3);
\end{tikzpicture}
\end{center}
%
The arrows above show the direction of influence; and the implication of $x\rightarrow y$ and $y \rightarrow z$ is that $x$ affects $z$ in two ways. First, it has (some) direct effect; and second, it can have indirect effect by influencing $y$, regulated by parameter $\rho$ as shown in \nameref{S3_Appendix}.

To make inferences about this mediated effect, we need more than one model. First, the regression of $z$ on $x$, i.e., the \emph{non-mediated} (or univariate) model which tells on the \emph{total} effect of $x$ on $z$; and second, the regression of $z$ on $x$ and $y$, i.e., the \emph{mediated} (or multivariate or full) model. We simulated the data as shown in \nameref{S3_Appendix} and fitted the two models -- non-mediated and mediated. In both cases, we compared mean-based and the whole-sample-based predictions, as shown in Fig~\ref{fig:pred_mediated_plots}.

\begin{figure}
\begin{center}
\includegraphics{mediate_preds.ggp.pdf}
\end{center}
\caption{{\bf Mean-based and whole-sample-based predictions for mediated effects.} These figures compare the mean-based and whole-sample-based predictions for the mediated (multivariate) and non-mediated (univariate) models. A: In the absence of mediator variable, both mean-based and whole-sample-based approaches predictors are identical since there is no any other additional (non-focal input variables) sources of bias. Consequently, for both approaches, the average predicted probability is very close to the observed proportion. B: When the mediator variable is included,  $x$ has no direct effect on $z$ and as a result the predictions do not align with the observations. However, the whole-sample-based approach still closely approximates the average proportion in the simulated data. The horizontal dashed lines are the respective average predictions and observed proportions. The vertical dashed black represents the center point, and the point at which it crosses the horizontal lines represents the expected ``perfect'' prediction at the model center. The grey points are the binned observations.}
\label{fig:pred_mediated_plots}
\end{figure}

From Fig~\ref{fig:pred_mediated_plots}A, we see what we would expect in the absence of additional sources of bias, even though in the simulation, the effect of $x$ on $z$ is mediated through $y$. This is the total effect of $x$, which only tells us the influence of $x$ on $z$. By ignoring $y$ in the model, we are still able to capture the effect of $x$ and closely match the observed values using both approaches. However, if there were additional non-focal predictors, we would expect to see the differences similar to those in Fig~\ref{fig:pred_bin_plots}. Including both $y$ and $x$ ``dilutes'' the direct effect of $x$ on $z$ and as a result, our predictions do not necessarily match the observed binned observations (see Fig~\ref{fig:pred_mediated_plots}B).

@@

Our simulations examples focused on simple linear and logistic models due their wide range of usage and application. These models also act as a starting point for building other complex models, including mixed effect models and models with categorical predictors. The logic for extending to more complex models, including other forms of nonlinear link functions is very straightforward. In fact the components needed for extension are the correct linear predictor and the inverse link function; everything else generalizes. In addition, our \proglang{R} package implementation already extends to and supports most of the nonlinear link functions and mixed model framework, including multivariate binary outcome models.

@@

Generating outcome predictions or predicted probabilities from simple and generalized linear (mixed) models is not only important but also, generating quantities which are consistent with the observed values should be of interest. However, many studies still report coefficient estimates from generalized models like probit, logistic, etc., \citep{hanmer2013behind}, which are subject of less clarity and complexities in interpretation.

The argument and results we present in this paper supports a greater need for a shift on focus on what and how to present predictions from generalized models. For example, the effect-styled confidence intervals could provide more clarity concerning the uncertainty due to the input variable of interests as opposed to the conventional way of incorporating everything.

@@
